---
title: "Custom-class-and-dplyr-compatibility"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Custom-class-and-dplyr-compatibility}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ExtendDataFrames)
```

## Extending Data Frames in R

R is a commonly used language for data science and statistical computing. Foundational to this is having data structures that allow manipulation of data with minimal effort and cognitive load. One of the most commonly required data structures is tabular data. This can be represented in R in a few ways, for example a matrix or a data frame. The data frame (class `data.frame`) is a flexible tabular data structure, as it can hold different data types (e.g. numbers, character strings, etc.) across different columns. This is in contrast to matrices -- which are arrays with dimensions -- and thus can only hold a single data type. 

```{r}
# data frame can hold heterogeneous data types across different columns
data.frame(a = c(1, 2, 3), b = c(4, 5, 6), c = c("a", "b", "c"))

# each column must be of the same type
df <- data.frame(a = c(1, 2, 3), b = c("4", 5, 6))
# be careful of the silent type conversion
df$a
df$b

mat <- matrix(1:9, nrow = 3, ncol = 3)
mat
mat[1, 1] <- "1"
# be careful of the silent type conversion
mat
```

Data frames can even be nested, cells can be data frames or lists.

[INSERT CODE EXAMPLE OF NESTED DF HERE]

It is therefore clear why data frames are so prevalent. However, they are not without limitations. They have a relatively basic printing method which can fload the R console when the number of columns or rows is large. They have useful methods (e.g., `summary()` and `str()`), but these might not be appropriate for certain types of tabular data. In these cases it is useful to utilise R's inheritance mechanisms (specifically S3 inheritance) to write extensions for R's `data.frame` class. In this case the data frame is the superclass and the new subclass extends it and inherits its methods (see Adv R[^1] for more details on S3 inheritance).

One of the most common extension of the data frame is the `tibble` from the {tibble} R package. Outlined in {tibble}'s 'Tibbles' vignette[^2], `tibble`s offer improvements in printing, subsetting and recycling rules. Another commonly used data frame extension is the `data.table` class from the {data.table} R package [^3]. This class is primarily designed to improve the performance (i.e. speed and efficiency of operations and storage) of working with tabular data in R. 

In the process of developing R software (most likely an R package), a new tabular data class that builds atop data frames can become beneficial. This blog post has two main sections: 
1) a brief overview of  the steps required to setup a class that extends data frames
2) guide to the technical aspects of class invariants and design and implementation decisions, and tidyverse compatibility 

### Writing custom data class

It is useful to write a class constructor function that can be called to create an object of your new class. The functions define below are a redacted version (for readability) of functions also defined in the package.

```r
birthdays <- function(x) {
	# the vector of classes is required for it to inherit from `data.frame`
  structure(x, class = c("birthdays", "data.frame"))
}
```

That's all that's needed to create a subclass of a data frame. However, although we've created the class we haven't given it any functionality and thus it will be identical to a data frame due to inheritance.

We can now write as many methods as we want. Here we will show two methods, one of which does not require writing a generic and the second that does.

```r
print.birthdays <- function(x, ...) {
  cat(
    sprintf(
      "A `birthdays` object with %s rows and %s cols",
      dim(x)[1], dim(x)[2]
    )
  )
  invisible(x)
}

birthdays_per_month <- function(x, ...) {
  UseMethod("birthdays_per_month")
}

birthdays_per_month.birthdays <- function(x, ...) {
  out <- table(lubridate::month(x$birthday))
  months <- c(
    "Jan", "Feb", "Mar", "Apr", "May", "Jun",
    "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"
  )
  names(out) <- months[as.numeric(names(out))]
  return(out)
}
```

Useful resources for the "Writing custom data class" section:

- [extending `tibbles` and their functionality](https://tibble.tidyverse.org/articles/extending.html)

### Design decision around class invariants

Moving onto the second section of the post, in which we discuss the design choices when creating and using S3 classes in R. ***Class invariants*** are members of your class that define it. In other words, without these elements your class does not fulfil its basic definition. It is therefore sensible to make sure that your class contains these elements at all times (or at least after operations have been applied to your class). In cases when the class object contains all the invariants normal service can be continued. However, in the case that an invariant is missing or modified to a non-conformist type (e.g. a date converted to a numer) a decision has to be made. Either the code can error, hopefully giving the user an informative message as to why their modification broke the object; alternatively, the subclass can be revoke and the superclass can be returned. In almost all cases the superclass (i.e. the base class being inherited from) is more general and won't have the same class invariant restrictions. 

For our example class, `birthdays`, the invariants are a column called `name` which must contain characters, and a column called `birthday` which must contain dates. The order of the rows and columns is not considered an invariant property, and having extra columns with other names and data types is also allowed. The number of rows is also not an invariant as we can have as many birthdays as we like in the data object.

Here we present both cases and considerations and technical details of both options. We'll demonstrate both of these cases with the subset function in R (subsetting uses a single square bracket for tabular data, `[`). First the fail-on-subsetting. Before we write the subsetting function it is useful to have a function that checks that an object of our class is valid, a so-called validator function.

```{r}
validate_birthdays <- function(x) {
  stopifnot(
    "input must contain 'name' and 'birthday' columns" =
      all(c("name", "birthday") %in% colnames(x)),
    "names must be a character" =
      is.character(x$name),
    "birthday must be a date" =
      lubridate::is.Date(x$birthday)
  )
  invisible(x)
}
```


This will return an error if the class is not valid (defined in terms of the class' invariants). 

Now we can show how to error if one of the invariants are removed during subsetting. 

```{r, eval=FALSE}
`[.birthdays` <- function(x) {
  validate_birthdays(NextMethod())
}

birthdays[, -1]
	Error in validate_birthdays(NextMethod()) :
	input must contain 'name' and 'birthday' columns
```

The second design option is the check whether the class is valid, and if not downgrade the class to the superclass, in our case a data frame. This is done by not only validating the object during subsetting but to check whether it is a valid class object, and then either ensuring all of the attributes of the subclass -- in our case `birthdays` -- are maintained, or attributes are stripped and only the attributes of the base superclass -- in our case `data.frame` -- are kept. 

___Important note: this section of the post relies heavily on code from___ https://github.com/DavisVaughan/2020-06-01_dplyr-vctrs-compat.

The four functions that are required to be added to ensure our class is correctly handled when invaliding it are: 

- `birthdays_reconstruct()`
- `birthdays_can_reconstruct()`
- `df_reconstruct()`
- `dplyr_reconstruct.birthdays()`

We'll tackle the first three first, and then move onto to the last one as this requires some extra steps. 

`birthdays_reconstruct()` is a function that contains an if-else statement to determine whether the returned object is a `birthdays` or `data.frame` object.

```{r}
birthdays_reconstruct <- function(x, to) {
  if (birthdays_can_reconstruct(x)) {
    df_reconstruct(x, to)
  } else {
    class(x) <- "data.frame"
    message("Removing crucial column in `<birthdays>` returning `<data.frame>`")
    x
  }
}
```

The if-else evaluation is controlled by `birthdays_can_reconstruct()`. This function determines whether after subsetting the object is a valid `<birthdays>` class. It does this by running the class validator, and if validation fails, by throwing an error, it returns `FALSE`, otherwise the function will return `TRUE`.

```{r}
birthdays_can_reconstruct <- function(x) {
  # check whether input is valid
  valid <- tryCatch(
    { validate_birthdays(x) },
    error = function(cnd) FALSE
  )

  # return boolean
  !isFALSE(valid)
}
```

The next function required is `df_reconstruct()`. This is called when the object is judged to be a valid `<birthdays>` object and simply copies the attributes over from the `<birthdays>` class to the object being subset.

```{r}
df_reconstruct <- function(x, to) {
  attrs <- attributes(to)
  attrs$names <- names(x)
  attrs$row.names <- .row_names_info(x, type = 0L)
  attributes(x) <- attrs
  x
}
```

The three functions defined for reconstruction can be added to a package with the subsetting function in order to subset `<birthdays>` objects and returning either `<birthdays>` objects if still valid, or data frames when invalidated. This design has the benefit that when conducting data exploration a user is not faced with an error, but can continue with a data frame, while being informed by the message printed to console in `birthdays_reconstruct()`.

```{r}
`[.birthdays` <- function(x, ...) {
  out <- NextMethod()
  birthdays_reconstruct(out, x)
}
```
[^1]: https://adv-r.hadley.nz/s3.html?q=inherita#s3-inheritance
[^2]: https://tibble.tidyverse.org/articles/tibble.html
[^3]: https://github.com/Rdatatable/data.table
