---
title: "ExtendDataFrames"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ExtendDataFrames}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ExtendDataFrames)
```

In the {ExtendDataFrames} package we need to create a class. For this example,
We have created the `<birthdays>` class. It is very simple, fundamentally the
class just needs two columns, with column names: `name` and `birthday`. The
names must be exact (i.e. case- and plural-sensitive), but we'll come back to
this aspect later. The role of the class -- and it's important that a well-designed 
class has a role, otherwise just using the build in data frame will 
probably suffice -- is to store the name of a person and their birthday. 

For a full explaination of the functions used in this vignette, look at the 
source code and see the other vignette.

We can create a `<birthdays>` object by calling the class constructor. First
we create a data frame that can be converted to a `<birthdays>` object and pass
that to the constructor, which is called `birthdays()`. In other R packages 
class constructors may be called `new_[class-name]()`, `build_[class-name]()`,
or similar.

```{r}
df <- data.frame(
  name = c("kevin", "stacey"),
  birthday = c(as.Date("2001-01-01"), as.Date("2002-01-02"))
)
birthdays <- birthdays(df)
birthdays
```

We can check whether our new object has been assigned the correct class 
attribute.

```{r}
class(birthdays)
attributes(birthdays)
```

The benefit of extending the data frame class is we inherit all of the methods
already written for data frames, and can overwrite (overload) these methods with
our own custom `<birthday>` methods. We have written a custom `print()`
(`print.birthdays()`) method and will use the inherited `summary()` and `str()`
methods.

```{r}
print(birthdays)
summary(birthdays)
str(birthdays)
```

The print method returns invisibly which we can check:

```{r}
tmp <- print(birthdays)
identical(tmp, birthdays)
```

We can also write custom methods that are not already defined, in other words
do not already have a defined generic function. Here we write a 
`birthdays_per_month()` method which does as it says on the tin.

```{r}
birthdays_per_month(birthdays)
```

The `<birthdays>` class is validated upon construction and can also be validated
interactively when working with it:

```{r}
validate_birthdays(birthdays)

validate_birthdays(birthdays[1, ])

try(validate_birthdays(birthdays[, 1]))

birthdays$age <-  seq_len(nrow(birthdays))
validate_birthdays(birthdays)
```

For a full explanation of the behaviour in the above code chunk see the other 
vignette.

Other methods for the class can also be added, for example:

```{r}
birthday_paradox(birthdays)
```

Another example of making a class with randomised names and birthdays:

```{r}
df <- data.frame(
  name = randomNames::randomNames(10),
  birthday = sample(
    x = seq.Date(
      from = as.Date("2020-01-01"),
      to = as.Date("2022-01-01"),
      by = 1),
    size = 10
  )
)
birthdays <- birthdays(df)
birthdays
as.data.frame(birthdays)
```
